\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-11-03 paper/main.tex
illustrative example walkthrough up through line 14

% ======== additional package not in SoftwareX template ========
\usepackage[colorlinks=true, urlcolor=blue, pdfborder={0 0 0}]{hyperref}
\usepackage{breakurl}
\usepackage[htt]{hyphenat}
%\usepackage{listings}
\usepackage{xcolor}
%\usepackage{todonotes}

@@

Within the Python~\cite{vanR95} community, external
packages that are published in the most popular
repositories~\cite{Pyth03, cond15} are associated with version numbers
and tags that allow users to guarantee they are installing
exactly the same code across different computing environments \cite{CoghStuf13}.
While it is \textit{possible} to manually install the intended
version of every dependency of a Python script or package, manually
tracking down those dependencies can impose a substantial burden on
the user and create room for mistakes and inconsistencies. Further, when dependency versions
are left unspecified, replicating the original computing environment
becomes difficult or impossible.

@@

>>>>

% ======== additional package not in SoftwareX template ========
\usepackage[colorlinks=true, urlcolor=blue, pdfborder={0 0 0}]{hyperref}
\usepackage{breakurl}
\usepackage[htt]{hyphenat}
%\usepackage{listings}
%\usepackage{xcolor}
%\usepackage{todonotes}
\usepackage{placeins}

@@

Within the Python~\cite{vanR95} community, external
packages that are published in the most popular
repositories~\cite{Pyth03, cond15} are associated with version numbers
and tags that allow users to guarantee they are installing
exactly the same code across different computing environments~\cite{CoghStuf13}.
While it is \textit{possible} to manually install the intended
version of every dependency of a Python script or package, manually
tracking down those dependencies can impose a substantial burden on
the user and create room for mistakes and inconsistencies. Further, when dependency versions
are left unspecified, replicating the original computing environment
becomes difficult or impossible.

@@

Line 6 then uses the \texttt{davos.config} object to enable \texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next two packages, \texttt{NumPy}~\cite{HarrEtal20} and \texttt{pandas}. Because these packages rely heavily on custom C data types, loading them may require restarting the notebook kernel if different versions were already imported during the same interpreter session (see Section~\ref{subsec:config}). This is unlikely to happen with \texttt{NumPy}, as it is loaded explicitly only once (by line 7) and is not required by any packages loaded before it (and therefore not implicitly imported when loading them). However, it's possible a future user could edit this notebook to add additional code before line 7, run the notebook's cells out of order, or even configure their IPython kernel to execute custom code on startup (e.g., as Google Colab notebooks do), any of which could potentially import a different, existing version of \texttt{NumPy} before \texttt{davos} smuggles the version specified in the onion comment. Enabling the \texttt{auto\_rerun} option before smuggling \texttt{NumPy} would simplify running the notebook in these scenarios, but the main benefit it affords comes when smuggling the \texttt{pandas} library.

The dataset that the user wants to analyze was provided as a \texttt{pandas.Panel} object, serialized with Python's built-in \texttt{pickle} module. The \texttt{pickle} protocol is a popular choice for data persistence in Python, allowing users to save, share, and load arbitrary objects, though with an important caveat: A ``pickled'' object's class definition must be importable from the same module as when it was saved in order to successfully ``unpickle'' (i.e., load and restore) it later. The \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the user's dataset can be loaded only if their notebook uses a \texttt{pandas} version released prior to that change. However, updates to the \texttt{pandas} library since then have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats (including in Jupyter notebooks).
%Recent updates to the \texttt{pandas} library have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats (including in Jupyter notebooks). However, the \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the dataset they are working with can only be loaded by a version released prior to that change.
\texttt{davos} makes it possible to do both within a single notebook, and with the \texttt{auto\_rerun} option, this process can be fully automated. The first time the notebook is run, lines 10--12 will install and load an older \texttt{pandas} version that defines the \texttt{Panel} class, read the dataset from its serialized format, convert it from a \texttt{Panel} to a multi-index \texttt{DataFrame}, and write its contents to a CSV file. The \texttt{smuggle} statement on line 14 then installs a newer \texttt{pandas} version (v1.3.5) for use in the analyses themselves. Because an earlier version of \texttt{pandas} was previously loaded (by line 10), and because the package's C extension modules were updated between the old and new versions, switching to the new version will require restarting the notebook kernel. With the \texttt{auto\_rerun} option enabled, \texttt{davos} will handle this automatically, and once the kernel is restarted, will re-run the notebook up through the current cell (as well as re-queue any other cells that were queued before restarting). This second time the notebook is run (as well as any future times),
